Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-16 paper/paper.tex
more progress

We extend the metric in Equation \eqref{eq:mmd} for the purpose of quantifying model shifts. Specifically, we introduce \textbf{Predicted Probability MMD (PP MMD)}: instead of applying Equation \eqref{eq:mmd} to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model \(M\). If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a meshgrid over a subspace of the entire feature space. For the latter approach we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension \(D\) grows.

@@

Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 10,000 individuals each. We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.

@@

Figure \ref{fig:syn} also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently. Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP. It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.

@@

\caption{Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points. Error bars indicate the standard deviation across folds.}\label{fig:syn}
\end{figure}

@@

The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large. Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

@@

\caption{Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points. Error bars indicate the standard deviation across folds.}\label{fig:real}
\end{figure}

@@

\caption{The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.}\label{fig:mitigate-results}
\end{figure}

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. A snapshot of the results is shown in Figure \ref{fig:mitigate-latent-results}.

@@

\caption{Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.}\label{fig:mitigate-latent-results}
\end{figure}

>>>>

We extend the metric in Equation \eqref{eq:mmd} for the purpose of quantifying model shifts. Specifically, we introduce \textbf{Predicted Probability MMD (PP MMD)}: instead of applying Equation \eqref{eq:mmd} to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model \(M\). If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a mesh grid over a subspace of the entire feature space. For the latter approach we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension \(D\) grows.

@@

Since the simulations involve generating counterfactuals for a significant proportion of the entire sample of individuals, we have randomly undersampled each dataset to yield balanced subsamples consisting of 2,500 individuals each. We have also standardized all explanatory features since our chosen classifiers are sensetive to scale.

@@

Figure \ref{fig:syn} also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently. Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP. It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy, that explicitly address this quantity, fail to work as expected.

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful\footnote{We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.}. Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \ref{fig:syn}. The same broadly holds for the Moons data, For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment. That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.

@@

\caption{Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of 1,000 points. Error bars indicate the standard deviation across folds.}\label{fig:syn}
\end{figure}

@@

The same is not true for model shifts shown the middle row of Figure \ref{fig:real}: with the exception of GMSC, the estimated PP MMD is statistically significant and large. Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

@@

\caption{Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid. Error bars indicate the standard deviation across folds.}\label{fig:real}
\end{figure}

@@

\caption{The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds.}\label{fig:mitigate-results}
\end{figure}

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. In Figure \ref{fig:mitigate-latent-results} the conventional Latent Space generator with \(\gamma=0.5\) serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

@@

\caption{Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds.}\label{fig:mitigate-latent-results}
\end{figure}

Finally, Figure \ref{fig:mitigate-real-world-results} shows the results for our real-world data. We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \ref{fig:mitigate-results}\footnote{Estimated domain shifts (not shown) were largely insubstantial, as in Figure \ref{fig:real} in the previous section.}. Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points. For the GMSC dataset we observe no notable differences.

{\centering \includegraphics[width=0.9\linewidth]{www/mitigation_real_world_results}

\caption{The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$. Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid. Error bars indicate the standard deviation across folds.}\label{fig:mitigate-real-world-results}
\end{figure}

### 2022-08-16 paper/sections/empirical_2.rmd
more progress

Figure \@ref(fig:syn) also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently. Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP. It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy that explicitly address this quantity fail to work as expected.

```{r syn, fig.cap="Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of 1,000 points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/synthetic_results.png")
```

@@

The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated grid-based PP MMD is statistically significant and large. Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

@@

```{r real, fig.cap="Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over a meshgrid of 10,000 points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/real_world_results.png")
```

>>>>

Figure \@ref(fig:syn) also indicates that the estimated effects are strongest for the simplest linear classifier, a pattern that we have observed fairly consistently. Conversely, there is virtually no difference in outcomes between the deep ensemble and the MLP. It is possible that the deep ensembles simply fail to capture predictive uncertainty well and hence counterfactual generators like Greedy, that explicitly address this quantity, fail to work as expected.

The findings for the other synthetic datasets are broadly consistent with the observations above. For the Moons data the same broad patterns emerge, although in this case it is less evident that Latent Space generators induce relatively smaller shifts. For the Circles data, it also appears at first sight that Latent Space search yields better results, but it turns out that in this case counterfactual search is simply largely unsuccessful^[We suspect that this in this case the generative model has failed to learn an accuracte representation of the data geenrating process.]. Model shifts and performance deterioration are also quantitatively quantitatively smaller than in what we can observe in Figure \@ref(fig:syn). The same broadly holds for the Moons data,  For the Linearly Separable data we also find substantial domain and model shifts, but no reduction in model performance.

Finally, it is also worth noting at this point that the observed dynamics and patterns are consistent throughout the course of the experiment. That is to say that we start observing shifts already after just a few rounds and these tends to increase proportionately for the different generators over the course of the experiment.

```{r syn, fig.cap="Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of 1,000 points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/synthetic_results.png")
```

@@

The same is not true for model shifts shown the middle row of Figure \@ref(fig:real): with the exception of GMSC, the estimated PP MMD is statistically significant and large. Here we find no evidence that Latent Space search helps to mitigate model shifts, as we did before for the synthetic data. Since these real-world datasets are arguably more complex than the synthetic data, the generative model can be expected to have a harder time at learning the data generating process and hence this increased difficult appears to affect the performance of REVISE/CLUE.

@@

```{r real, fig.cap="Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/real_world_results.png")
```

### 2022-08-16 paper/sections/methodology_2.rmd
more progress

We extend the metric in Equation \@ref(eq:mmd) for the purpose of quantifying model shifts. Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying Equation \@ref(eq:mmd) to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$. If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a meshgrid over a subspace of the entire feature space. For the latter approach we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.

>>>>

We extend the metric in Equation \@ref(eq:mmd) for the purpose of quantifying model shifts. Specifically, we introduce **Predicted Probability MMD (PP MMD)**: instead of applying Equation \@ref(eq:mmd) to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model $M$. If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a mesh grid over a subspace of the entire feature space. For the latter approach we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension $D$ grows.

### 2022-08-16 paper/sections/mitigation.rmd
more progress

```{r mitigate-results, fig.cap="The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\\gamma=0.5$. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_results.png")
```

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. A snapshot of the results is shown in Figure \@ref(fig:mitigate-latent-results).

```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_latent_results.png")
```

>>>>

```{r mitigate-results, fig.cap="The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\\gamma=0.5$. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_results.png")
```

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

```{r mitigate-latent-results, fig.cap="Combinining various mitigation strategies with Latent Space search. Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a mesh grid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_synthetic_latent_results.png")
```

Finally, Figure \@ref(fig:mitigate-real-world-results) shows the results for our real-world data. We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.]. Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points. For the GMSC dataset we observe no notable differences.

```{r mitigate-real-world-results, fig.cap="The differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\\gamma=0.5$. Results for deep ensemble using real-world datasets. The shown model MMD (PP MMD) was computed over actual samples, rather than a mesh grid. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/mitigation_real_world_results.png")
```
